---
title: "Basic Exploratory Plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{basic_plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(dlptools)
```


Most of the plots here related to exploring aspects of a DLP run to help learn how well it may have worked or not!

A good resource of exploring aspects of DLP are these confluence pages:

* [DLP Library Analysis Basics](https://molonc.atlassian.net/wiki/spaces/SOFT/pages/2702082064/DLP+Library+Analysis+Basics)
* [Comparison of SCP and Mondrian](https://molonc.atlassian.net/wiki/spaces/SOFT/pages/2473459715/DLP+pipeline+comparison+scp+vs+mondrian)
* [Explanation of Metrics Columns](https://molonc.atlassian.net/wiki/spaces/SOFT/pages/2459041795)



## Chip Layout Plots

The first thing we usually want to check, is how the chip is laid out:

```{r, fig.width=8, fig.height=5}
# just some example data from a recent run with lots of experimental conditions
ex_mets <- vroom::vroom(
  "data/ex_metrics.tsv.gz",
  show_col_types = FALSE
)

dlptools::chip_plot(ex_mets, "experimental_condition")
```


<br>
<br>

Then often we want to look for spatial effects across the chip.

Such as, quality:

```{r, fig.width=8, fig.height=5}
dlptools::chip_plot(ex_mets, "quality")
```

<br>
<br>

or perhaps total reads:

```{r, fig.width=8, fig.height=5}
dlptools::chip_plot(ex_mets, "total_reads")
```


<br>
<br>

## GC Correction Plots

Another plot that can be useful is to look at the GC correction for a cell, to see if the values really fall on integers for CNs, or if maybe something seems off about the multiplier HMMcopy chose for the cell.

```{r, fig.width=8, fig.height=5}
# standard reads table, really anything with GC, modal_curve, multiplier, and
# cell_id columns should work.
ex_reads <- vroom::vroom(
  "data/example_full_reads.tsv.gz",
  show_col_types = FALSE
)

dlptools::gc_plot(
  ex_reads,
  cellid = "AT21352-A144173A-R03-C32",
  # plot_choice = "both" # the default option
)
```

<br>

Or you can just plot one:

```{r, fig.width=6, fig.height=4}
dlptools::gc_plot(
  ex_reads,
  cellid = "AT21352-A144173A-R03-C32",
  plot_choice = "corrected" # or plot_choice = "raw"
)
```

<br>
<br>
<br>

## Cell Copy Profiles

Sometimes you want to show a plot of a cell, or handful of cells, with their copy values and the overlaid state calls.

The Signals package has [plotCNprofile](https://shahcompbio.github.io/signals/reference/plotCNprofile.html) which has lots of features.

But if you want a quick and dirt dlptools version, you can do:

```{r, fig.width=10, fig.height=5}
dlptools::cell_cn_profile(
  ex_reads,
  cell_ids = unique(ex_reads$cell_id)[1:2],
  # high copy numbers can obscure the plot, so you can set:
  # pseduo_log_y = TRUE
  # and that might help. Or see function help for more ideas
)
```

<br>
<br>
<br>

If needed, can also change what the y-axis is to other bin-based options:

```{r, fig.width=10, fig.height=5}
dlptools::cell_cn_profile(
  ex_reads,
  cell_ids = unique(ex_reads$cell_id)[1:2],
  yaxis = "state"
)
```
