---
title: "Basic Exploratory Plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{basic_plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dlptools)
```


Most of the plots here related to exploring aspects of a DLP run to help learn how well it may have worked or not!

A good resource of exploring aspects of DLP are these confluence pages:

* [DLP Library Analysis Basics](https://molonc.atlassian.net/wiki/spaces/SOFT/pages/2702082064/DLP+Library+Analysis+Basics)
* [Comparison of SCP and Mondrian](https://molonc.atlassian.net/wiki/spaces/SOFT/pages/2473459715/DLP+pipeline+comparison+scp+vs+mondrian)
* [Explanation of Metrics Columns](https://molonc.atlassian.net/wiki/spaces/SOFT/pages/2459041795)



## Chip Layout Plots

The first thing we usually want to check, is how the chip is laid out:

```{r}
# just some example data from a recent run with lots of experimental conditions
ex_mets <- vroom::vroom("data/ex_metrics.tsv.gz")

dlptools::chip_plot(ex_mets, "experimental_condition")
```


<br>
<br>

Then often we want to look for spatial effects across the chip.

Such as, quality:

```{r}
dlptools::chip_plot(ex_mets, "quality")
```

<br>
<br>

or perhaps total reads:

```{r}
dlptools::chip_plot(ex_mets, "total_reads")
```


<br>
<br>
