---
title: "Feature Extraction for CN Signatures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{feature_extraction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(dlptools)
```

## Features Defined by Other Publications

dlptools offers two externally defined feature extraction types:

1. Wu et al. https://www.biorxiv.org/content/10.1101/2025.03.02.641098v1
2. Wang et al. https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1009557

The Wang et al features are just a thin wrapper around [sigminer](https://github.com/ShixiangWang/sigminer?tab=readme-ov-file) functions. Check out that package for more tools and actual signature extraction.

<br>

As well, dlptools offers functions to extract the standard "**process based**" features, defines/inspired by:

* Macintyre et al. 2018
* Maclachlan et al. 2021
* Drews et al. 2022
* and others

<br>

These features include:

* segment size
* copy number change points
* breakpoint counts per arm or MB window
* length of oscillating copy number changes

Often, extracting these features is a starting point, and they are further summarized into categories. These categories are either set by hand, or informed by mixture models.

Some publications also include the actual CN of the segments (in categories) as a feature, but I haven't included a function for that...presumably you can pull that out of your data.

<br>
<br>

A likely starting point of this exercise might be some read bin data, or some segment data. It's probably good that it's been cleaned up to your liking.

Perhaps the workflow would look something like this:

```{r, warning=FALSE, message = FALSE}
reads_df <- vroom::vroom("data/example_full_reads.tsv.gz")

segs_df <- reads_df |>
  dplyr::filter(!chr %in% c("X", "Y")) |>
  dlptools::mark_mask_regions() |>
  dlptools::mark_bins_overlapping_centromeres(padding = 3e6) |>
  dplyr::filter(
    gc != -1 & map > 0.99 & !overlaps_centro & !mask
  ) |>
  dlptools::add_missing_bins_for_cells() |>
  dlptools::fill_state_from_neighbours() |>
  dlptools::reads_to_segs() |>
  dplyr::filter(
    seg_width > 1.5e6
  ) |>
  dlptools::segs_to_reads() |>
  dlptools::add_missing_bins_for_cells() |>
  dlptools::fill_state_from_neighbours() |>
  dlptools::reads_to_segs()


dplyr::slice_head(segs_df, n = 5)
```
 


<br>
<br>

### Wu et al.

For extracting Wu et al. features:

```{r}
dlptools::extract_wu_features(
  segs_df = segs_df
) |>
  dplyr::slice_head(n = 5)
```
<br>
<br>

There are other options, like annotating the input dataframe, or returning a matrix for subsequent signature extraction:

```{r}
feat_mtx <- dlptools::extract_wu_features(
  segs_df = segs_df,
  return_matrix = TRUE
)

feat_mtx[1:4, 1:4]

# alternatively, there is a generic dlptools::make_cellid_matrix() than can
# pivot out and make the matrix if you already have the count dataframe and
# don't want to redo the feature extraction
```

<br>

see `?dlptools::extract_wu_features` for more.

<br>
<br>
<br>

### Wang et al., aka Sigminer

Alternatively, we can get Wang et al. features. Though if going this route, it would be a good idea to look more at [sigminer](https://github.com/ShixiangWang/sigminer?tab=readme-ov-file) itself:

```{r, warning=FALSE, message = FALSE}
sig_feats <- extract_sigminer_wang_features(segs_df)

sig_feats[1:4, 1:4]
```

<br>
<br>
<br>

## Process-Based Features

1. segment lengths. 

This is as simple as it sounds, and the function is unnecessary. These lengths would need to be subsequently categorized somehow.

```{r}
dlptools::extract_segment_sizes(
  segs_df = segs_df
) |>
  dplyr::slice_head(n = 10)
```

<br>

2. Copy number change points

These are the values of the change in copy number from segment to segment. E.g., 2-5 would yield a 3.

See `?dlptools::extract_changepoint` for an important discussion of the `first_seg_correction` parameter.

```{r}
dlptools::extract_changepoint(
  segs_df = segs_df
) |>
  dplyr::slice_head(n = 10)
```

<br>

3. Extracting number of breakpoints per chromosome arm

A breakpoint is a change in the copy number.

```{r}
dlptools::extract_breakpoints(
  segs_df = segs_df,
  scope = "arms"
) |>
  dplyr::slice_head(n = 10)

# there is an alias of: dlptools::extract_bp_per_arm()
```

<br>

4. Extract breakpoints per MB window

Typically, publications have used 10Mb.

```{r}
dlptools::extract_breakpoints(
  segs_df = segs_df,
  scope = "windows"
  # window_size = 10e6 # default
) |>
  dplyr::slice_head(n = 10)

# there is an alias of: dlptools::extract_bp_per_window()
```

<br>

5. lengths of copy number oscillations


Overall, this is a bit of a funny feature. But basically, chromosomes are assessed in 3-segment sets to determine if it counts as an oscillation. If there are more than one oscillations in a row, they are added up. If the 3-segment set is not an oscillation, it gets a 0.

<br>

* 1-2-3 is not an oscillation, length 0
* 2-3-2 is an oscillation of length 1
* 2-3-2-3 is an oscillation of length 2
* 2-3-2-3-5 is an oscillation of length 2 followed by a 0 for the non-oscillating component
* 1-2-3-4-5 has no oscillations, and returns 0 0 0 for the 3 possible 3-set segments

<br>

I have dissected the code for various publications and packages that count this feature and they all work differently, some have outright bugs, and others lack some internal consistency in how they assess 0s. So, please read `?dlptools::extract_oscillations` for a long discussion of how exactly this is working.

<br>

Overall, 0s are typically abundant, and should maybe be considered for removal...? I don't know, they just tend to overwhelm. Again, all publications using this feature have done so differently, so it's the wild west. Do what feels right.

```{r}
dlptools::extract_oscillations(
  segs_df = segs_df
  # important arguments are:
  # middle_bound = 2 # default
  # ends_bound = 0 # default
  # see function documentation for what these are
) |>
  dplyr::slice_head(n = 10)
```

<br>

You can also play directly with `dlptools::count_oscillations` on vectors to see how the inernals of this function are working:

```{r}
dlptools::count_oscillations(1:3)

dlptools::count_oscillations(c(2, 3, 2))

dlptools::count_oscillations(c(2, 3, 2, 3))

dlptools::count_oscillations(c(2, 3, 2, 3, 8, 5, 4, 5))

# most other publications/package do not care about the middle value
dlptools::count_oscillations(
  c(5, 1000, 5),
  middle_bound = Inf
)

# some publications would allow this as counting as an oscillation. Where the
# first and last segments are allowed to be within a certain distance of one
# another, and the middle value doesn't matter.
dlptools::count_oscillations(
  c(5, 1000, 6),
  middle_bound = Inf,
  ends_bound = 1
)
```


<br>
<br>
<br>

## Other vaguely feature like things

These could be considered features, maybe only within certain contexts. Maybe they are just more useful as descriptions of the data as a whole.

### ploidy informed CN change

CN change relative to ploidy might be interesting to have a feature. Here we can mark 3 types of events for segments:

1. CN is a gain relative to mode ploidy of sample/cell
2. CN is a loss relative to mode ploidy of sample/cell
3. CN is a match to mode ploidy

Employs the function dlptools::mode_ploidy() to calculate mode ploidy of each sample/cell.

```{r}
dlptools::extract_ploidy_cn_feature(
  segs_df,
  # with option just to annotate the input df
  # annotate_input = TRUE
  # or, alternatively, return a count matrix
  # return_matrix = TRUE
) |>
  dplyr::slice_head(n = 6)
```

<br>

### Segment Span Types

It might be useful to know where segments sit on chromosomes and how much of a chromosome they cover. We can label segments based on 5 categories:

1. telomere bound (`telo-bound`) - segment touches a telomere
2. centromere bound (`centro-bound`) - segment touches or crosses the centromere
3. arm (`arm`) - segment spans a whole are (*with conditions)
4. whole chromosome (`whole-chrom`) - segment spans the entire chromosome (*with conditions)
5. intersitial (`inter`) - occuring within the chromosome, not touching the centromere, telomeres, and not big enough to be an entire arm.

See `?dlptools::mark_segs_chromosome_span` for more details about this marking, as there are various options to tweak these definitions, such as:

* how much of an arm a CN needs to cover to be considered an "arm-level" event
* how much of a chromosome a segment needs to cover to be a "whole-chromosome" event
* how close a seg needs to be to a telo/centromere to be considered "touching"

We can mark a dataframe with the information:

```{r, message=FALSE, warning=FALSE}
dlptools::mark_segs_chromosome_span(
  segs_df,
  # see ?dlptools::mark_segs_chromosome_span for more args
) |>
  dplyr::relocate(
    # key added columns
    seg_span_event, telo_bound, centro_bound, spans_chrom, spans_arm
  ) |>
  dplyr::slice_head(n = 5)
```

<br>
<br>

More useful in the "feature" context, is to directly extract and count them up. This function bascially just calls the one above, with counting added:

```{r, message=FALSE, warning=FALSE}
dlptools::extract_segment_position_feature(
  segs_df
  # with option just to annotate the input df
  # annotate_input = TRUE
  # or, alternatively, return a count matrix
  # return_matrix = TRUE
) |>
  dplyr::slice_head(n = 6)
```


## Feature Combinations...?

With some of these feature sets, we can combine them into one monster feature set. Is this a good idea? A bad idea? Is the output even interpretable? Are all ov these even useful? You'll have to decide that.

The feature sets of:

1. `dlptools::extract_wu_features`
2. `dlptools::extract_ploidy_cn_feature`
3. `dlptools::extract_segment_position_feature`

are all "atomic", in that they describe aspects of individual segments.

First, we can collect labels all all features for the segments:

```{r, message=FALSE, warning=FALSE}
feat_anno_segs <- segs_df |>
  dlptools::extract_wu_features(annotate_input = TRUE) |>
  dlptools::extract_ploidy_cn_feature(annotate_input = TRUE) |>
  dlptools::extract_segment_position_feature(annotate_input = TRUE) |>
  dplyr::select(
    cell_id, chr, start, end,
    # these are then the feature columns
    seg_bin, cn_bin, seg_change, seg_shape, cn_v_ploidy, seg_span_event
  )


dplyr::slice_head(feat_anno_segs, n = 5)
```


Then, we can just count each category and leave the counts per cell per feature as our feature set for signature extraction. This would yield a 21 feature set (6 core feature types with: 5, 3, 3, 2, 3, 5 categories across them)

<!-- Seeing NA in AA/BB for this dataset...track down! -->


```{r, message=FALSE, warning=FALSE}
feat_cols <- c(
  "seg_bin", "cn_bin", "seg_change", "seg_shape", "cn_v_ploidy",
  "seg_span_event"
)

total_counts <- purrr::map(feat_cols, \(feat_col) {
  feat_anno_segs |>
    dplyr::group_by(cell_id, dplyr::pick(feat_col), .drop = FALSE) |>
    dplyr::count() |>
    dplyr::mutate(
      feature = feat_col
    ) |>
    dplyr::rename(category = feat_col) |>
    dplyr::relocate(cell_id, feature, category)
}) |>
  purrr::list_rbind()
```

```{r, echo=FALSE, fig.width=8, fig.height=4}
ggplot(
  dplyr::filter(total_counts, cell_id %in% unique(total_counts$cell_id)
  [1:4]),
  aes(x = category, y = n, fill = feature)
) +
  geom_col(show.legend = FALSE) +
  facet_grid(
    cell_id ~ feature,
    scales = "free",
    axes = "all_x",
    switch = "x"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    strip.placement = "outside"
  )
```

<br>
<br>

We could opt to cross all of the features into a "mega-feature-set". But, this would be:

* 4 features (containing 3, 5, 2, 3 categories across them) with from `dlptools::extract_wu_features`
* 1 feature with 3 features categories from `dlptools::extract_ploidy_cn_feature`
* 1 feature with 5 feature categories from `dlptools::extract_segment_position_feature`

```{r}
all_feat_labels <- list(
  levels(feat_anno_segs$seg_bin),
  levels(feat_anno_segs$cn_bin),
  levels(feat_anno_segs$seg_change),
  levels(feat_anno_segs$seg_shape),
  levels(feat_anno_segs$cn_v_ploidy),
  levels(feat_anno_segs$seg_span_event)
)

crossed_labels <- expand.grid(all_feat_labels)

crossed_labels[1:4, ]
```


```{r}
nrow(crossed_labels)
```


making for a cross feature set with 1350 possible categories. That might be too detailed, and you should check if most categories are 0 for your data, indicating useless features. 

<br>

Also, probably want to ask if crossing the category actually makes sense and adds biological information. For example, here, you can't even have a small sized (< 5Mb) arm event (smallest arm is >10 Mb), so crossing size and chromosome location doesn't actually lead to fully crossed features or any gain in information (and these two are somewhat redundant). Or, does it matter if the segment shape parameter is a ploidy-gain or a ploidy-loss? Does knowing that distinguish processes or teach you anything biologically? (and that's only two parameters being crossed).

<br>

But, forging ahead bravely:

```{r}
# collect all possible feature labels

crossed_counts <- feat_anno_segs |>
  dplyr::group_by(
    # sample ID column
    cell_id,
    # feature columns
    seg_bin, cn_bin, seg_change, seg_shape, cn_v_ploidy, seg_span_event,
    .drop = FALSE
  ) |>
  dplyr::count() |>
  dplyr::mutate(
    cross_feat = stringr::str_c(
      seg_bin, cn_bin, seg_change, seg_shape, cn_v_ploidy, seg_span_event,
      sep = ":"
    )
  )
```

```{r, echo=FALSE, fig.width=8, fig.height=4}
ggplot(
  dplyr::filter(crossed_counts, cell_id %in% unique(total_counts$cell_id)
  [1:4]),
  aes(x = cross_feat, y = n)
) +
  geom_col() +
  facet_grid(cell_id ~ ., axes = "all_x") +
  theme_classic() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
  ) +
  xlab("Crossed Feature")
```


<br>
<br>

Or, you could do some sort of combination of cross and uncrossed? I don't know. Again, this whole signature field is the wild west and people do all kinds of things.

```{r, message=FALSE, warning=FALSE}
# count the Wu features as crossed: 90 feature categories
wu_crossed_counts <- dlptools::extract_wu_features(segs_df) |>
  dplyr::select(cell_id, feat_cat, n) |>
  dplyr::mutate(
    feature_group = "Wu-crossed"
  )

# do ploidy_cn_change and segment placement as uncrossed?
ploidy_counts <- dlptools::extract_ploidy_cn_feature(segs_df) |>
  dplyr::rename(feat_cat = cn_v_ploidy) |>
  dplyr::mutate(
    feature_group = "ploidy"
  )


seg_pos_counts <- dlptools::extract_segment_position_feature(segs_df) |>
  dplyr::rename(feat_cat = seg_span_event) |>
  dplyr::mutate(
    feature_group = "position"
  )

all_feat_counts <- purrr::list_rbind(list(
  wu_crossed_counts, ploidy_counts, seg_pos_counts
)) |>
  dplyr::arrange(cell_id, feature_group) |>
  dplyr::mutate(
    feature_group = factor(feature_group, levels = unique(feature_group))
  )
```


```{r, echo=FALSE, fig.width=8, fig.height=4}
ggplot2::ggplot(
  dplyr::filter(all_feat_counts, cell_id %in% unique(total_counts$cell_id)
  [1:4]),
  ggplot2::aes(x = feat_cat, y = n, fill = feature_group)
) +
  ggplot2::geom_col(show.legend = FALSE) +
  ggplot2::facet_grid(
    cell_id ~ feature_group,
    axes = "all_x", scales = "free_x", space = "free"
  ) +
  ggplot2::scale_fill_manual(values = dlptools::GEN_PLOT_COLS) +
  ggplot2::theme_classic() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
  ) +
  ggplot2::xlab("Mixed Features")
```


<br>
<br>
<br>
