---
title: "Feature Extraction for CN Signatures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{feature_extraction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dlptools)
```

dlptools offers two feature extraction types:

1. Wu et al. https://www.biorxiv.org/content/10.1101/2025.03.02.641098v1
2. Wang et al. https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1009557

The Wang et al features are just a thin wrapper around [sigminer](https://github.com/ShixiangWang/sigminer?tab=readme-ov-file) functions. Check out that package for more tools and actual signature extraction.

<br>
<br>

A likely starting point of this exercise might be some read bin data, or some segment data. It's probably good that it's been cleaned up to your liking.

Perhaps the workflow would look something like this:

```{r, warning=FALSE, message = FALSE}
reads_df <- vroom::vroom("data/example_full_reads.tsv.gz")

segs_df <- reads_df |>
  dlptools::mark_mask_regions() |>
  dlptools::mark_bins_overlapping_centromeres(padding = 3e6) |>
  dplyr::filter(
    gc != -1 & map > 0.99 & !overlaps_centro & !mask
  ) |>
  dlptools::add_missing_bins_for_cells() |>
  dlptools::fill_state_from_neighbours() |>
  dlptools::reads_to_segs() |>
  dplyr::filter(
    seg_width > 1.5e7
  ) |>
  dlptools::segs_to_reads() |>
  dlptools::add_missing_bins_for_cells() |>
  dlptools::fill_state_from_neighbours() |>
  dlptools::reads_to_segs()


dplyr::slice_head(segs_df, n = 5)
```
 


<br>
<br>

For extracting Wu et al. features:

```{r}
dlptools::extract_wu_features(
  segs_df = segs_df
) |>
  dplyr::slice_head(n = 5)
```
<br>
<br>

There are other options, like annotating the input dataframe, or returning a matrix for subsequent signature extraction:

```{r}
feat_mtx <- dlptools::extract_wu_features(
  segs_df = segs_df,
  return_matrix = TRUE
)

feat_mtx[1:4, 1:4]
```

<br>

see `?dlptools::extract_wu_features` for more.

<br>
<br>
<br>

Alternatively, we can get Wang et al. features. Though if going this route, it would be a good idea to look more at sigminer itself:

```{r, warning=FALSE, message = FALSE}
sig_feats <- extract_sigminer_wang_features(segs_df)

sig_feats[1:4, 1:4]
```

<br>
<br>
<br>
