---
title: "Feature Extraction for CN Signatures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{feature_extraction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(dlptools)
```

## Feature Extraction

`dlptools` has a series of functions to extract a variety of features. These could be used for subsequent signature extraction, or be use to quantify differences between samples.

On offer are the "classic" process based features, inspired by 

* [Macintyre et al. 2018](https://www.nature.com/articles/s41588-018-0179-8)
* [Maclachlan et al. 2021](https://www.nature.com/articles/s41467-021-25469-8)
* [Drews et al. 2022](https://www.nature.com/articles/s41586-022-04789-9)
* and others

These include:

1. segment size
2. breakpoints per arm
3. breakpoints per window
4. changepoints: change in CN state between adjacent segments
5. length of oscillating chains
6. CN state (though there is debate if this should be used at all)

<br>

The functions for these (outlined below) will extract the observed values for each of these features, and offers to summarise counts in pre-defined categories. These categories are informed by the publications above, but designed to be accommodating for DLP data, which has a more coarse resolution than most of those publications.

<br>

Additionally, there are functions to extract two externally defined feature extraction types:

1. Wu et al. https://www.biorxiv.org/content/10.1101/2025.03.02.641098v1
2. Wang et al. https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1009557

The Wang et al features are just a thin wrapper around [sigminer](https://github.com/ShixiangWang/sigminer?tab=readme-ov-file) functions. Check out that package for more tools and actual signature extraction.


<br>
<br>

A likely starting point of this exercise might be some read bin data, or some segment data. It's probably good that it's been cleaned up to your liking.

Perhaps the workflow would look something like this:

```{r, warning=FALSE, message = FALSE}
reads_df <- vroom::vroom("data/example_full_reads.tsv.gz")

segs_df <- reads_df |>
  dplyr::filter(!chr %in% c("X", "Y")) |>
  dlptools::mark_mask_regions() |>
  dlptools::mark_bins_overlapping_centromeres(padding = 3e6) |>
  dplyr::filter(
    gc != -1 & map > 0.99 & !overlaps_centro & !mask
  ) |>
  dlptools::add_missing_bins_for_cells() |>
  dlptools::fill_state_from_neighbours() |>
  dlptools::reads_to_segs() |>
  dplyr::filter(
    seg_width > 1.5e6
  ) |>
  dlptools::segs_to_reads() |>
  dlptools::add_missing_bins_for_cells() |>
  dlptools::fill_state_from_neighbours() |>
  dlptools::reads_to_segs()


dplyr::slice_head(segs_df, n = 5)
```
 
<br>
<br>
<br>
<br>

## Process-Based Features

For the 6 functions in this section, each by default return the observed values. But, each function also has an option to `return = counts`, which will use pre-defined categories.

<br>

1. segment lengths. 

This is as simple as it sounds, and the function is unnecessary. These lengths would need to be subsequently categorized somehow.

```{r}
dlptools::extract_segment_sizes(
  segs_df = segs_df
) |>
  dplyr::slice_head(n = 10)
```

<br>

alternatively you can specify the `return = counts` to get counts in pre-defined categories. This is an argument available for all of these process based functions.

```{r}
dlptools::extract_segment_sizes(
  segs_df = segs_df,
  return = "counts"
) |>
  dplyr::slice_head(n = 10)
```

categories are: < 5mb, 5-10 Mb, 20-50 Mb, 50-100 Mb, and 100+Mb.


<br>

2. Copy number change points

These are the values of the change in copy number from segment to segment. E.g., 2-5 would yield a 3.

See `?dlptools::extract_changepoint` for an important discussion of the `first_seg_correction` parameter.

```{r}
dlptools::extract_changepoint(
  segs_df = segs_df
  # or returning the counts
  # return = 'counts'
) |>
  dplyr::slice_head(n = 10)
```

Predefined count categories are changepoint = 1, 2, 3, 4, 5+

<br>

3. Extracting number of breakpoints per chromosome arm

A breakpoint is a change in the copy number.

```{r}
dlptools::extract_breakpoints(
  segs_df = segs_df,
  scope = "arms"
  # or returning the counts
  # return = 'counts'
) |>
  dplyr::slice_head(n = 10)

# there is an alias of: dlptools::extract_bp_per_arm()
```

Predefined count categories are: breakpoint = 0, 1, 2, 3, 4, 5, 6+

<br>

4. Extract breakpoints per MB window

Typically, publications have used 10Mb.

```{r}
dlptools::extract_breakpoints(
  segs_df = segs_df,
  scope = "windows"
  # window_size = 10e6 # default
  # or returning the counts
  # return = 'counts'
) |>
  dplyr::slice_head(n = 10)

# there is an alias of: dlptools::extract_bp_per_window()
```

Predefined count categories are: breakpoint = 0, 1, 2, 3+

<br>

5. lengths of copy number oscillations


Overall, this is a bit of a funny feature. But basically, chromosomes are assessed in 3-segment sets to determine if it counts as an oscillation. If there are more than one oscillations in a row, they are added up. If the 3-segment set is not an oscillation, it gets a 0.

<br>

* 1-2-3 is not an oscillation, length 0
* 2-3-2 is an oscillation of length 1
* 2-3-2-3 is an oscillation of length 2
* 2-3-2-3-5 is an oscillation of length 2 followed by a 0 for the non-oscillating component
* 1-2-3-4-5 has no oscillations, and returns 0 0 0 for the 3 possible 3-set segments

<br>

I have dissected the code for various publications and packages that count this feature and they all work differently, some have outright bugs, and others lack some internal consistency in how they assess 0s. So, please read `?dlptools::extract_oscillations` for a long discussion of how exactly this is working.

<br>

Overall, 0s are typically abundant, and should maybe be considered for removal...? I don't know, they just tend to overwhelm. Again, all publications using this feature have done so differently, so it's the wild west. Do what feels right.

```{r}
dlptools::extract_oscillations(
  segs_df = segs_df
  # or returning the counts
  # return = 'counts'

  # other important arguments are:
  # middle_bound = 2 # default
  # ends_bound = 0 # default
  # see function documentation for what these are
) |>
  dplyr::slice_head(n = 10)
```

Predefined count categories are: Oscillation chains of length: 0, 1-3, 4-9, 10+.

E.g.,

```{r}
dlptools::extract_oscillations(
  segs_df = segs_df,
  return = "counts"
) |>
  dplyr::slice_head(n = 10)
```


<br>

You can also play directly with `dlptools::count_oscillations` on vectors to see how the inernals of this function are working:

```{r}
dlptools::count_oscillations(1:3)

dlptools::count_oscillations(c(2, 3, 2))

dlptools::count_oscillations(c(2, 3, 2, 3))

dlptools::count_oscillations(c(2, 3, 2, 3, 8, 5, 4, 5))

# most other publications/package do not care about the middle value
dlptools::count_oscillations(
  c(5, 1000, 5),
  middle_bound = Inf
)

# some publications would allow this as counting as an oscillation. Where the
# first and last segments are allowed to be within a certain distance of one
# another, and the middle value doesn't matter.
dlptools::count_oscillations(
  c(5, 1000, 6),
  middle_bound = Inf,
  ends_bound = 1
)
```

<br>

6. CN States

And for sake of completeness, also offered is CN. By default, function just pulls the CN state column from your dataframe. 

Really only slightly useful if you want to use the pre-defined categories of:

CN = 1, 2, 3, 4, 5, 6+

```{r}
dlptools::extract_cn(
  segs_df,
  return = "counts"
) |>
  dplyr::slice_head(n = 10)
```

<br>
<br>

If you are happy with all of the default categories for counting, you can call the wrapper to get them all at once:

```{r, warning=FALSE, message=FALSE, error=FALSE}
# if you set up a future::plan(), this will run in parallel. See ?dlptools::count_default_process_feats

# could be something like
# future::plan(future::multicore, workers = 10)

dlptools::count_default_process_feats(
  segs_df
  # CN not included by default, use this to include it
  # with_cn=TRUE
) |>
  dplyr::slice_head(n = 10)
```


<br>
<br>

### Wu et al.

For extracting Wu et al. features:

```{r}
dlptools::extract_wu_features(
  segs_df = segs_df
) |>
  dplyr::slice_head(n = 5)
```
<br>
<br>

There are other options, like annotating the input dataframe, or returning a matrix for subsequent signature extraction:

```{r}
feat_mtx <- dlptools::extract_wu_features(
  segs_df = segs_df,
  return_matrix = TRUE
)

feat_mtx[1:4, 1:4]

# alternatively, there is a generic dlptools::make_cellid_matrix() than can
# pivot out and make the matrix if you already have the count dataframe and
# don't want to redo the feature extraction
```

<br>

see `?dlptools::extract_wu_features` for more.

<br>
<br>
<br>

### Wang et al., aka Sigminer

Alternatively, we can get Wang et al. features. Though if going this route, it would be a good idea to look more at [sigminer](https://github.com/ShixiangWang/sigminer?tab=readme-ov-file) itself:

```{r, warning=FALSE, message = FALSE}
sig_feats <- extract_sigminer_wang_features(segs_df)

sig_feats[1:4, 1:4]
```


<br>
<br>
<br>

## Other vaguely feature like things

These could be considered features, maybe only within certain contexts. But, maybe they are just more useful as descriptions of the data as a whole.

### ploidy informed CN change

CN change relative to ploidy might be interesting to have a feature. Here we can mark 3 types of events for segments:

1. CN is a gain relative to mode ploidy of sample/cell
2. CN is a loss relative to mode ploidy of sample/cell
3. CN is a match to mode ploidy

Employs the function dlptools::mode_ploidy() to calculate mode ploidy of each sample/cell.

```{r}
dlptools::extract_ploidy_cn_feature(
  segs_df,
  # with option just to annotate the input df
  # annotate_input = TRUE
  # or, alternatively, return a count matrix
  # return_matrix = TRUE
) |>
  dplyr::slice_head(n = 6)
```

<br>

### Segment Span Types

It might be useful to know where segments sit on chromosomes and how much of a chromosome they cover. We can label segments based on 5 categories:

1. telomere bound (`telo-bound`) - segment touches a telomere
2. centromere bound (`centro-bound`) - segment touches or crosses the centromere
3. arm (`arm`) - segment spans a whole are (*with conditions)
4. whole chromosome (`whole-chrom`) - segment spans the entire chromosome (*with conditions)
5. intersitial (`inter`) - occuring within the chromosome, not touching the centromere, telomeres, and not big enough to be an entire arm.

See `?dlptools::mark_segs_chromosome_span` for more details about this marking, as there are various options to tweak these definitions, such as:

* how much of an arm a CN needs to cover to be considered an "arm-level" event
* how much of a chromosome a segment needs to cover to be a "whole-chromosome" event
* how close a seg needs to be to a telo/centromere to be considered "touching"

We can mark a dataframe with the information:

```{r, message=FALSE, warning=FALSE}
dlptools::mark_segs_chromosome_span(
  segs_df,
  # see ?dlptools::mark_segs_chromosome_span for more args
) |>
  dplyr::relocate(
    # key added columns
    seg_span_event, telo_bound, centro_bound, spans_chrom, spans_arm
  ) |>
  dplyr::slice_head(n = 5)
```

<br>
<br>

More useful in the "feature" context, is to directly extract and count them up. This function bascially just calls the one above, with counting added:

```{r, message=FALSE, warning=FALSE}
dlptools::extract_segment_position_feature(
  segs_df
  # with option just to annotate the input df
  # annotate_input = TRUE
  # or, alternatively, return a count matrix
  # return_matrix = TRUE
) |>
  dplyr::slice_head(n = 6)
```



<br>
<br>
<br>
