---
title: "Pairwise Differences Between Cells"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pairwise-differences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dlptools)
```


The functions here were inspired by a paper called: *Genomic evolution and cellular states of whole-genome doubling in ovarian cancer.* By Weiner et al. 2025, some folks at MSKCC.

The basic idea is to count the numbers of bins that are different in state between two cells, divided by the number of bins considered. Then we can do things like fit a beta distribution to identify outliers, or do other clustering to find groups of similar/different cells, etc.

These functions will make that a bit easier.

First some toy data:

```{r}
reads_df <- vroom::vroom("data/example_reads.tsv.gz", show_col_types = FALSE)

# only using a few cells to demonstrate functions below
targ_cells <- unique(reads_df$cell_id)[1:40]

reads_df <- dplyr::filter(reads_df, cell_id %in% targ_cells)

# standard reads dataframe of bin based state calls
dplyr::slice_head(reads_df, n = 5)
```

<br>
<br>

Now, we can measure pairwise distances between all cells. This involves:

1. generating all pairwise cell comparisons
2. aligning bins for each pair
3. collapsing runs of bins based on *matched* and *unmatched* states between the cell pair -- aka segmenting on state matches
4. filtering small runs (segments) of matches or non-matches
5. re-binning the runs
6. counting number of non-matching bins, divided by total number of bins still considered
7. presenting the results for each cell and it's distances to all other cells

<br>

Warning, this function is *slow*, and the number of pairwise comparisons grows with more cells. E.g., 100 cells is 4950 comparisons and takes a couple minutes. There is definitely room for improvement here.

Parallelizing will make your life a bit better. Internally, this function uses `furrr`, you just need to set up a plan.


```{r, message=FALSE, warning=FALSE, eval=FALSE}
# first set up a parallel plan, here with 4 cores being used
# future::plan(future::multisession, workers = 4)

pairwise_diffs <- dlptools::pairwise_bin_difference(
  reads_df,
  # min_seg_length = 2.5e6 # see function documentation. It's the minimum
  # length of a stretch of matching/unmatching bins to consider.
)

dplyr::slice_head(pairwise_diffs, n = 5)
```

```{r, message=FALSE, warning=FALSE, echo=FALSE}
pairwise_diffs <- vroom::vroom(
  "data/stored_pairwise_diffs.tsv",
  show_col_types = FALSE
)

dplyr::slice_head(pairwise_diffs, n = 5)
```



which is each cell (`index cell`) compared to each other cell (`comp_cell`) and some information on the proportion of bins different (`prop_diff`).

<br>
<br>
<br>

We can then find the nearest neighbour of each cell:

```{r}
(nn_cells <- dlptools::find_nearest_neighbours(pairwise_diffs))
```


<br>
<br>
<br>

---

**Side Note**

The function `dlptools::pairwise_bin_difference()` has a `targ_cells` parameter. Leaving it empty, the default, compares all cells in a pairwise manner.

This will compare the specified cell or cells against all others:

```{r, message=FALSE, warning=FALSE}
dlptools::pairwise_bin_difference(
  reads_df,
  targ_cells = unique(reads_df$cell_id)[1:2]
)
```


<br>
<br>

Alternatively, if you want to just compare some set of cells against themselves, best to just filter upfront:

```{r, message=FALSE, warning=FALSE}
reads_df |>
  dplyr::filter(cell_id %in% unique(reads_df$cell_id)[1:3]) |>
  dlptools::pairwise_bin_difference()
```

<br>

---

<br>
<br>

We can then combine the pairwise differences and nearest neighbours to find outlier cells by fitting a beta distribution to the latter, and finding outlier pairwise comparisons:


```{r}
(outlier_cells <- dlptools::find_outlier_cells(pairwise_diffs, nn_cells))
```

<br>
<br>
<br>

There is a simple plot that can be made to visualize these distances:

```{r, fig.height=4, fig.width=6}
dlptools::plot_nnd_outlier_cells(
  pairwise_diffs, nn_cells, outlier_cells
)
```


<br>
<br>
<br>


With a bit of work, we can re-orient the cell distances matrix to contain each cell compared against all others. The original function performs non-redundant comparisons: A-B, A-C, B-C. 

For various reasons, we might also want to see C-A and C-B, even though it's redundant.

This is actually happening internally in several of the above functions.

This dataframe can get big, as it's now every cell compared to every other cell.

```{r}
# first, we need to know all of the cells we want to do this for. It could
# be all cells, or some subset.

# the nearest neighbours frame is all input cells, so could use that.

cell_focused_df <- purrr::map_dfr(
  # might want to consider furrr::future_map_dfr
  unique(nn_cells$index_cell),
  \(cell) {
    dlptools::make_cell_focused_matrix(pairwise_diffs, cell)
  }
)

dplyr::slice_head(cell_focused_df, n = 5)
```


Now every cell will be present in the `index_cell` column.

<br>
<br>
<br>
