---
title: "Pairwise Differences Cetween Cells"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pairwise-differences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dlptools)
```


The functions here were inspired by a paper called: *Genomic evolution and cellular states of whole-genome doubling in ovarian cancer.* By Weiner et al. 2025, some folks at MSKCC.

The basic idea is to count the numbers of bins that are different in state between two cells, divided by the number of bins considered. Then we can do things like fit a beta distribution to identify outliers, or do other clustering to find groups of similar/different cells, etc.

These functions will make that a bit easier.

First some toy data:

```{r}
reads_df <- vroom::vroom("data/example_reads.tsv.gz", show_col_types = FALSE)

# only using a few cells to demonstrate functions below
targ_cells <- unique(reads_df$cell_id)[1:50]

reads_df <- dplyr::filter(reads_df, cell_id %in% targ_cells)

# standard reads dataframe of bin based state calls
dplyr::slice_head(reads_df, n = 5)
```

<br>
<br>

Now, we can measure pairwise distances between all cells. This involves:

1. generating all pairwise cell comparisons
2. aligning bins for each pair
3. collapsing runs of bins based on *matched* and *unmatched* states between the cell pair -- aka segmenting on state matches
4. filtering small runs (segments) of matches or non-matches
5. re-binning the runs
6. counting number of non-matching bins, divided by total number of bins still considered
7. presenting the results for each cell and it's distances to all other cells

<br>

Warning, this function is *slow*, and the number of pairwise comparisons grows with more cells. E.g., 100 cells is 4950 comparisons and takes a couple minutes. There is definitely room for improvement here.

Parallelizing will make your life a bit better. Internally, this function uses `furrr`, you just need to set up a plan.

```{r}
# first set up a parallel plan, here with 4 cores being used
future::plan(future::multicore, workers = 4)

pairwise_diffs <- dlptools::pairwise_bin_difference(
  reads_df,
  # min_seg_length = 2.5e6 # see function documentation. It's the minimum
  # length of a stretch of matching/unmatching bins to consider.
)

dplyr::slice_head(pairwise_diffs, n = 5)
```


which is each cell (`index cell`) compared to each other cell (`comp_cell`) and some information on the proportion of bins different (`prop_diff`).

The `nearest_neighbour` column is a boolean where the nearest neighbour of each cell is marked as TRUE.

<br>
<br>
<br>

---

**Side Note**

The function `dlptools::pairwise_bin_difference()` has a `cells` parameter. Leaving it empty, the default, compares all cells in a pairwise manner.

This will compare this one cell against all others:

```{r, eval=FALSE}
dlptools::pairwise_bin_difference(
  reads_df,
  cells = "some_cell_id"
)
```


Alternatively, specifying two or more cells will just compared the specified cells to each other:

```{r, eval=FALSE}
dlptools::pairwise_bin_difference(
  reads_df,
  cells = c("cell_id_one", "cell_id_two", "cell_id_three")
)
```

---

<br>
<br>

The nearest neighbour of each cell is marked in the dataframe returned above. This function now takes that information, fits a beta distribution to those nearest neighbours, then returns outlier cells based on that distribution:

```{r}
outlier_cells <- dlptools::find_outlier_cells(pairwise_diffs)

dplyr::slice_head(outlier_cells, n = 5)
```

<br>
<br>
<br>

There is a simple plot that can be made to visualize these distances:

```{r, fig.height=4, fig.width=6}
dlptools::plot_nnd_outlier_cells(
  pairwise_diffs, outlier_cells
)
```


<br>
<br>
<br>



You can also convert those distance measurements into a pairwise distance matrix:

```{r}
pairs_mtx <- dlptools::convert_dists_to_pairwise(pairwise_diffs)

pairs_mtx[1:4, 1:4]
```

<br>

or if you know you only want this, it can be returned with the function above:

```{r, eval=FALSE}
dlptools::pairwise_bin_difference(
  reads_df,
  return_pairs_matrix = TRUE # this option here
)
```


<br>
<br>

A matrix like this can then be useful for other analyses, like if you wanted to do some clustering:

```{r, fig.height=5, fig.width=7}
pairs_mtx |>
  as.dist() |>
  hclust() |>
  plot(labels = FALSE)
```

