---
title: "Various Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Various Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```


```{r}
library(dplyr)
library(vroom)
library(dlptools)
```


# Various Functions dlptools


## Masking Bad Regions

Masking regions that are bad for DLP, mostly the consequence of low mappability. 

```{r}
# assuming some sort of reads or segments DF with: chr, start, end
ex_reads <- vroom::vroom("data/example_reads.tsv.gz")

ex_reads <- dlptools::mark_mask_regions(ex_reads)

# adds a boolean `mask` column
ex_reads |>
  dplyr::select(cell_id, chr, start, end, mask) |>
  dplyr::slice_head(n = 5)
```


The default masking file is one constructed by Daniel Lai, and can be viewed at the [package source](https://github.com/molonc/dlptools/tree/dev/inst/extdata) or by loading:

```{r}
vroom::vroom(system.file("extdata", "blacklist_2023.07.17.txt", package = "dlptools"))
```


## Marking Centromere Locations

Centromeres are problematic for DLP data, the state calls are often inflated and highly inconsistent between adjacent bins. It's a messy region that we often want to omit. The default masking file (above) pretty much captures these regions, but sometimes we may want to be more specific. 

Alternatively, sometimes we want to specify windows around centromeres to include in filtering too.

The files loaded to mark centromeres come from [UCSC](https://hgdownload.soe.ucsc.edu/downloads.html) cytoband files of hg19 and hg38 releases.

```{r}
# easiest way is to just explictly mark whether bins are in centromeres
dlptools::mark_bins_overlapping_centromeres(
  reads_df = ex_reads,
  padding = 3e6, # e.g., specifying 3 Mb on each side of centromere
  # default padding is 0
  version = "hg19", # default, don't neeed to specify. Alternative is "hg38"
) |>
  dplyr::count(cell_id, overlaps_centro) |>
  dplyr::slice_head(n = 6)
```


Alternatively, this can be done in parts:

```{r}
# e.g., just adding the centromere information by chromosome
dlptools::add_centromere_locations(
  reads_df = ex_reads,
  version = "hg19" # default, don't need to specify. Alternative is "hg38"
) |>
  dplyr::select(
    cell_id, chr, centro_start, centro_end, start_p, end_p, start_q, end_q
  )

# internally this function is used to retrieve chromosome locations
# dlptools::load_ucsg_cenrtomeres()
```


## Information From Cell IDs

This is assuming our standard DLP cell ids, e.g., `AT23998-A138956A-R03-C34`.


```{r}
# single cell id
dlptools::sample_from_cell("AT23998-A138956A-R03-C34")

# single library ID
dlptools::library_from_cell("AT23998-A138956A-R03-C34")

# Also a generic function for either
# dlptools::pull_info_from_cell_id("AT23998-A138956A-R03-C34", 'sample')
# dlptools::pull_info_from_cell_id("AT23998-A138956A-R03-C34", 'library')


# multiple cell ids:
dlptools::sample_from_cell(ex_reads$cell_id[1:5])

# or library
dlptools::library_from_cell(ex_reads$cell_id[1:5])


# more useful it using it on your reads data frame
# extracting sample id and library id and inserting into the dataframe
ex_reads |>
  dplyr::mutate(
    sample_id = dlptools::sample_from_cell(cell_id),
    library_id = dlptools::library_from_cell(cell_id)
  ) |>
  dplyr::slice_sample(n = 5)
```


## Inferring Missing Bin Data

In the course of DLP analyses, we often filter bins, e.g., the masking regions explained above, bins within centromeres, bins with poor mappability or GC correction. Also, some methods we use (Signals, medicc), might drop data they don't like.

But that leaves gaps in the data, and limits the length of segments that can be inferred.

While the relative merit can be debated, one possible solution to this is to fill in the missing state data from an adjacent bin. I.e., if the bin on chromosome 1 from 1.5-2.0Mb is missing, use the state data from 1.0-1.5Mb to fill in the missing data for that bin.

These functions will help you do that.

First lets set up some filtered data:

```{r}
raw_reads <- vroom::vroom("data/example_reads.tsv.gz")

filt_reads <- raw_reads |>
  dlptools::mark_mask_regions() |>
  dlptools::mark_bins_overlapping_centromeres(padding = 3e6) |>
  dplyr::filter(
    gc != -1 & map > 0.99 & !overlaps_centro & !mask
  )


tibble::tibble(
  raw_bins = dplyr::distinct(raw_reads, chr, start, end) |> nrow(),
  filtered_bins = dplyr::distinct(filt_reads, chr, start, end) |> nrow()
)
```

<br>
<br>

Then we can insert missing bins. This will create NAs for all data columns except cell_id, chr, start, end, unless otherwise specified. Only cell level data should be specified to be carried through.

```{r}
# optional e.g., we can add some other fake cell level meta data to carry
# through for missing bins
filt_reads$library <- "fake_library"

filt_with_missing <- dlptools::add_missing_bins_for_cells(
  filt_reads,
  version = "hg19", # default, don't need to specify
  bin_size = 5e5, # default, don't need to specify
  # OPTIONAL! can also specify other columns to carry through for cells
  cell_metadata_cols = c("library")
)

filt_with_missing |>
  dplyr::relocate(library, cell_id) |>
  dplyr::slice_head(n = 10)
```

NAs are inserted for missing bins.

<br>
<br>

Then we can actually fill in those bins with their neighbours, and specify any  of the columns you want filled:

```{r}
dlptools::fill_state_from_neighbours(
  filt_with_missing,
  cols_to_fill = c("state", "gc") # default is "state", don't need to specify
) |>
  dplyr::slice_head(n = 10)
```


Of course, it could all be done in a pipe:

```{r, eval=FALSE}
raw_reads |>
  dlptools::mark_mask_regions() |>
  dlptools::mark_bins_overlapping_centromeres(padding = 3e6) |>
  dplyr::filter(
    gc != -1 & map > 0.99 & !overlaps_centro & !mask
  ) |>
  dlptools::add_missing_bins_for_cells() |>
  dlptools::fill_state_from_neighbours()
```


Functions being used internally include:

```{r}
create_expected_bins(
  version = "hg19", # default, or hg38
  bin_size = 5e5 # default
) |>
  dplyr::slice_head(n = 5)
```

or expanding any integer into bins:

```{r}
dlptools::expand_length_to_bins(10, bin_size = 5)
```


And chromosome lengths used are coming from USCS `chromInfo` files:

```{r}
dlptools::load_chrom_info_file(
  version = "hg19" # default, or "hg38"
) |>
  dplyr::slice_head(n = 5)
```



## Reads to Segments

Grouping read bins into contiguous segments (e.g. post filtering read bins, etc.).

```{r}
ex_segs <- dlptools::reads_to_segs(ex_reads)

# this is now runs of adjacent read bins with identical states collapesed
# into a single bin. Of course, bins are no longer of equal size.
ex_segs[1:4, ]
```


*warning:* this function will leave some unexpected gaps when dataframes have been filtered and bins removed. Inspect carfully if you have dropped bins from your dataframe.

## Segments to Reads

Split segments back into bins of some size.

```{r}
rebinned_reads <- dlptools::segs_to_reads(
  ex_segs,
  # bin_size = 5e5 # default, but can change
)

rebinned_reads[1:4, ]
```

And recovers our original reads:

```{r}
nrow(rebinned_reads) == nrow(ex_reads)
```

Useful for things like dropping small segs from data:

```{r}
filt_reads <- ex_reads |>
  dlptools::reads_to_segs() |>
  dplyr::filter(seg_width > 1.5e6) |>
  dlptools::segs_to_reads()
# some columns lost in the tansform due to combining reads to segs. Unclear how
# to carry all columns of information.

c(filt_reads = nrow(filt_reads), input_reads = nrow(ex_reads))
```



## Long to Wide Reads (or segments)

Some functions require read state information to be in wide format vs long, with cell_ids as rows and chr_start_end as columns, and the states as cells.

```{r}
ex_reads_w <- dlptools::convert_long_reads_to_wide(ex_reads)

ex_reads_w[1:4, 1:4]
```



<br>
<br>
<br>

## Basic Plots

This plot is a simplified alternative to the methods described in [the heatmaps vignette](https://molonc.github.io/dlptools/articles/heatmaps.html). There are no additions, like trees and annotations, but works for a variety of quick inspections.

```{r, out.width="75%", out.height="75%"}
dlptools::basic_tile_plot(
  # just filtering to make the plot smaller for this demonstration
  dplyr::filter(ex_reads, chr %in% c(7:9))
)
```

To help with plotting, a variety of commonly use color palettes are available:

```{r}
# standard state colors
dlptools::CNV_COLOURS

# typically used allele specific colors
dlptools::ASCN_COLORS

# typically used phase colors
dlptools::ASCN_PHASE_COLORS

# typically BAF scale is a circlize::colorRamp2 spanning
# standard green-grey-purple used for ASCN colors
# dlptools::BAF_COLORS
```

## Data Importing

If you've followed a semi-standard approach for downloading such as this:

```bash
# bash
cd where/to/save/my_dlp/

ticket="SC-8382"
azcopy copy https://singlecellresults.blob.core.windows.net/results/${ticket}/results/annotation/ ${ticket} --recursive
azcopy copy https://singlecellresults.blob.core.windows.net/results/${ticket}/results/hmmcopy/ ${ticket} --recursive

```

which produces a directory with this sort of structure:

```bash
├── SC-8382
│   ├── annotation
│   │   ├── metrics.csv.gz
│   │   ├── #[...other files ...]
│   └── hmmcopy
│   │   ├── #[...other files ...]
│       ├── reads.csv.gz
│       ├── segments.csv.gz
├── SC-8408
│   ├── annotation
│   │   ├── metrics.csv.gz
│   │   ├── #[...other files ...]
│   └── hmmcopy
│   │   ├── #[...other files ...]
│       ├── reads.csv.gz
│       ├── segments.csv.gz
├── SC-8650
```

then there are some functions that can help with loading data from these into a consistent dataframe:

```{r, eval=FALSE}
dlp_dir <- "/where/to/save/my_dlp/"

# loading metrics
metrics <- dlptools::import_dlp_files(dlp_dir, "metrics")

# loading reads
reads_df <- dlptools::import_dlp_files(dlp_dir, "reads")

# loading segments...but generally this isn't a great idea
# segs_df <- dlptools::import_dlp_files(dlp_dir, "segments")
```



## Other

Phylogenetic trees made by Stika take some formatting before they can be plotted:

```{r, eval=FALSE}
dlptools::format_sitka_tree()
```

this function drops `locus` tips and removes the `cell_` part of cell id names on tips. This way, the trees can be aligned to cell ids in the heatmaps.
