---
title: "Various Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Various Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```


```{r, warning=FALSE, message=FALSE, echo=FALSE}
library(dlptools)
```


# Various Functions dlptools


## Masking Bad Regions

Masking regions that are bad for DLP, mostly the consequence of low mappability. 

```{r}
# assuming some sort of reads or segments DF with: chr, start, end
ex_reads <- vroom::vroom("data/example_reads.tsv.gz")

ex_reads <- dlptools::mark_mask_regions(ex_reads)

# adds a boolean `mask` column
ex_reads |>
  dplyr::select(cell_id, chr, start, end, mask) |>
  dplyr::slice_head(n = 5)
```


The default masking file is one constructed by Daniel Lai, and can be viewed at the [package source](https://github.com/molonc/dlptools/tree/dev/inst/extdata) or by loading:

```{r}
vroom::vroom(
  system.file("extdata", "blacklist_2023.07.17.txt", package = "dlptools")
)
```

<br>
<br>

## Marking Centromere and Telomere Locations

Centromeres are problematic for DLP data, the state calls are often inflated and highly inconsistent between adjacent bins. It's a messy region that we often want to omit. The default masking file (above) pretty much captures these regions, but sometimes we may want to be more specific. 

Alternatively, sometimes we want to specify windows around centromeres to include in filtering too.

The files loaded to mark centromeres come from [UCSC](https://hgdownload.soe.ucsc.edu/downloads.html) cytoband files of hg19 and hg38 releases.

```{r}
# easiest way is to just explictly mark whether bins are in centromeres
dlptools::mark_bins_overlapping_centromeres(
  reads_df = ex_reads,
  padding = 3e6, # e.g., specifying 3 Mb on each side of centromere
  # default padding is 0
  version = "hg19", # default, don't neeed to specify. Alternative is "hg38"
) |>
  dplyr::count(cell_id, overlaps_centro) |>
  dplyr::slice_head(n = 6)
```


Alternatively, this can be done in parts:

```{r}
# e.g., just adding the centromere information by chromosome
dlptools::add_centromere_locations(
  cn_df = ex_reads,
  version = "hg19" # default, don't need to specify. Alternative is "hg38"
) |>
  dplyr::select(
    cell_id, chr, centro_start, centro_end, start_p, end_p, start_q, end_q
  )

# internally this function is used to retrieve chromosome locations
# dlptools::load_ucsc_centromeres()
```


Also available is telomere marking, but see `?dlptools::import_telos_file` for why this isn't the most in-depth or accurate thing:

```{r}
dlptools::add_telomere_positions(
  ex_reads
  # version = "hg19" is the default
) |>
  dplyr::relocate(
    # added columns
    telostart_p, teloend_p, telostart_q, teloend_q
  ) |>
  dplyr::slice_head(n = 5)
```

<br>
<br>

## Information From Cell IDs

This is assuming our standard DLP cell ids, e.g., `AT23998-A138956A-R03-C34`.


```{r}
# single cell id
dlptools::sample_from_cell("AT23998-A138956A-R03-C34")

# single library ID
dlptools::library_from_cell("AT23998-A138956A-R03-C34")

# Also a generic function for either
# dlptools::pull_info_from_cell_id("AT23998-A138956A-R03-C34", 'sample')
# dlptools::pull_info_from_cell_id("AT23998-A138956A-R03-C34", 'library')


# multiple cell ids:
dlptools::sample_from_cell(ex_reads$cell_id[1:5])

# or library
dlptools::library_from_cell(ex_reads$cell_id[1:5])


# more useful it using it on your reads data frame
# extracting sample id and library id and inserting into the dataframe
ex_reads |>
  dplyr::mutate(
    sample_id = dlptools::sample_from_cell(cell_id),
    library_id = dlptools::library_from_cell(cell_id)
  ) |>
  dplyr::slice_sample(n = 5)
```

<br>
<br>

## Inferring Missing Bin Data

In the course of DLP analyses, we often filter bins, e.g., the masking regions explained above, bins within centromeres, bins with poor mappability or GC correction. Also, some methods we use (Signals, medicc), might drop data they don't like.

But that leaves gaps in the data, and limits the length of segments that can be inferred.

While the relative merit can be debated, one possible solution to this is to fill in the missing state data from an adjacent bin. I.e., if the bin on chromosome 1 from 1.5-2.0Mb is missing, use the state data from 1.0-1.5Mb to fill in the missing data for that bin.

These functions will help you do that.

First lets set up some filtered data:

```{r}
raw_reads <- vroom::vroom("data/example_reads.tsv.gz")

filt_reads <- raw_reads |>
  dlptools::mark_mask_regions() |>
  dlptools::mark_bins_overlapping_centromeres(padding = 3e6) |>
  dplyr::filter(
    gc != -1 & map > 0.99 & !overlaps_centro & !mask
  )


tibble::tibble(
  raw_bins = dplyr::distinct(raw_reads, chr, start, end) |> nrow(),
  filtered_bins = dplyr::distinct(filt_reads, chr, start, end) |> nrow()
)
```

<br>
<br>

Then we can insert missing bins. This will create NAs for all data columns except cell_id, chr, start, end, unless otherwise specified. Only cell level data should be specified to be carried through.

```{r}
# optional e.g., we can add some other fake cell level meta data to carry
# through for missing bins
filt_reads$library <- "fake_library"

filt_with_missing <- dlptools::add_missing_bins_for_cells(
  filt_reads,
  version = "hg19", # default, don't need to specify
  bin_size = 5e5, # default, don't need to specify
  # OPTIONAL! can also specify other columns to carry through for cells
  cell_metadata_cols = c("library")
)

filt_with_missing |>
  dplyr::relocate(library, cell_id) |>
  dplyr::slice_head(n = 10)
```

NAs are inserted for missing bins.

<br>
<br>

Then we can actually fill in those bins with their neighbours, and specify any  of the columns you want filled:

```{r}
dlptools::fill_state_from_neighbours(
  filt_with_missing,
  cols_to_fill = c("state", "gc") # default is "state", don't need to specify
) |>
  dplyr::slice_head(n = 10)
```


Of course, it could all be done in a pipe:

```{r, eval=FALSE}
raw_reads |>
  dlptools::mark_mask_regions() |>
  dlptools::mark_bins_overlapping_centromeres(padding = 3e6) |>
  dplyr::filter(
    gc != -1 & map > 0.99 & !overlaps_centro & !mask
  ) |>
  dlptools::add_missing_bins_for_cells() |>
  dlptools::fill_state_from_neighbours()
```


Functions being used internally include:

```{r}
create_expected_bins(
  version = "hg19", # default, or hg38
  bin_size = 5e5 # default
) |>
  dplyr::slice_head(n = 5)
```

or expanding any integer into bins:

```{r}
dlptools::expand_length_to_bins(10, bin_size = 5)
```


And chromosome lengths used are coming from USCS `chromInfo` files:

```{r}
dlptools::load_chrom_info_file(
  version = "hg19" # default, or "hg38"
) |>
  dplyr::slice_head(n = 5)
```


<br>
<br>

## Reads to Segments

Grouping read bins into contiguous segments (e.g. post filtering read bins, etc.).

```{r}
ex_segs <- dlptools::reads_to_segs(ex_reads)

# this is now runs of adjacent read bins with identical states collapesed
# into a single bin. Of course, bins are no longer of equal size.
ex_segs[1:4, ]
```


*warning:* this function will leave some unexpected gaps when dataframes have been filtered and bins removed. Inspect carfully if you have dropped bins from your dataframe.

<br>
<br>

## Segments to Reads

Split segments back into bins of some size.

```{r}
rebinned_reads <- dlptools::segs_to_reads(
  ex_segs,
  # bin_size = 5e5 # default, but can change
)

rebinned_reads[1:4, ]
```

And recovers our original reads:

```{r}
nrow(rebinned_reads) == nrow(ex_reads)
```

Useful for things like dropping small segs from data:

```{r}
filt_reads <- ex_reads |>
  dlptools::reads_to_segs() |>
  dplyr::filter(seg_width > 1.5e6) |>
  dlptools::segs_to_reads()
# some columns lost in the tansform due to combining reads to segs. Unclear how
# to carry all columns of information.

c(filt_reads = nrow(filt_reads), input_reads = nrow(ex_reads))
```

<br>
<br>

## Cut Segments

Segments span different intervals for each cell, but some tools need those aligned (i.e., the same spans -- start to end -- must be present for each sample). You could break the segments back into 500 kb bins, but that is generally excessive. Instead, we can break segments into the smallest consistent intervals across all samples.

See the help of `?dlptools::cut_equal_segments` for details.

```{r}
dlptools::cut_equal_segments(ex_segs) |>
  dplyr::arrange(chr, start) |>
  dplyr::slice_head(n = 5)
```

Each segment now starts and ends at the same spot. When a long segment is broken, the copy number information is the same for each portion, nothing about the state calls changes. The longer segments are just made shorter based on the smallest segment across cells.

vs what was before:

```{r}
ex_segs |>
  dplyr::arrange(chr, start) |>
  dplyr::slice_head(n = 5)
```

<br>
<br>

## Where Segments Occur

Useful to know is where segments occur relative to chromosome features like the centromere and telomeres.

There are a few key arguments to this function, like setting distances for how close segments need to be to telomeres and centromeres to be consider "bound", and minimum spans of arms/chromosomes. 

Please see `?dlptools::mark_segs_chromosome_span` for more information on how this works:

```{r}
seg_locations <- dlptools::mark_segs_chromosome_span(
  ex_segs,
  # version = "hg19" assumed, but hg38 possible too.
)

dplyr::relocate(
  seg_locations,
  # key added columns, more present too
  seg_span_event, telo_bound, centro_bound, spans_chrom, spans_arm
)
```


Visually, this is what we're doing:

```{r, echo = FALSE, fig.width=8, fig.height=6, fig.cap="Labels of CN segments on chromosomes. Vertical lines indicate locations of telomeres and centromeres."}
chr_attrs <- dplyr::left_join(
  dlptools::load_ucsc_centromeres(),
  dlptools::import_telos_file(),
  by = dplyr::join_by(chrom == chr)
) |>
  dplyr::rename(
    chr = chrom
  ) |>
  dplyr::select(chr, centro_start, centro_end, dplyr::contains("telo")) |>
  tidyr::pivot_longer(cols = -chr) |>
  dplyr::mutate(
    plot_name = dplyr::case_when(
      stringr::str_detect(name, "start") ~ "start",
      stringr::str_detect(name, "end") ~ "end",
    ),
    section_id = dplyr::case_when(
      stringr::str_detect(name, "centro") ~ "centro",
      stringr::str_detect(name, "telo.*p") ~ "telo-p",
      stringr::str_detect(name, "telo.*q") ~ "telo-q"
    )
  ) |>
  dplyr::select(chr, section_id, plot_name, value)


targ_chrs <- c(1, 2, 8, 13, 14)

some_cells <- unique(seg_locations$cell_id)[1:5]

seg_locations$chr <- factor(
  seg_locations$chr,
  levels = gtools::mixedsort(unique(seg_locations$chr))
)

ggplot2::ggplot(
  dplyr::filter(seg_locations, cell_id %in% some_cells & chr %in% targ_chrs)
) +
  ggplot2::geom_vline(
    # add genome features. Hm, maybe just do as vlines
    data = dplyr::filter(chr_attrs, chr %in% targ_chrs),
    # aes(xmin=start, xmax=end, ymin=-1, ymax=-2, fill=section_id)
    ggplot2::aes(xintercept = value),
    color = "black"
  ) +
  ggplot2::geom_segment(
    ggplot2::aes(x = start, xend = end, y = state, color = seg_span_event),
    linewidth = 3
  ) +
  ggplot2::facet_grid(
    cell_id ~ chr,
    scales = "free",
    space = "free_x",
    switch = "x",
    axes = "all_x",
  ) +
  ggplot2::theme_classic() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    strip.placement = "outside",
    strip.background = ggplot2::element_blank()
  ) +
  ggplot2::scale_color_manual(values = dlptools::GEN_PLOT_COLS)
```



<br>
<br>

## Ploidy of samples

We can get an idea of the ploidy of a sample by taking the mean of the copy numbers of the segments. However, it's probably a good idea

* to weight by the size of segment, so large segments contribute more the ploidy than small segments
* calculate means for each chromosome, and then across all chromosomes so that each chromosome contributes equally.

This function does that:

```{r}
dlptools::weighted_ploidy(ex_segs) |>
  dplyr::slice_head(n = 5)
```

<br>

Alternative, if you want integers, is to get the mode ploidy. The should only be done for evenly sized bins, or the mode wouldn't make much sense.

```{r}
dlptools::mode_ploidy(
  ex_reads,
  sample_col = "cell_id"
) |>
  dplyr::slice_head(n = 5)
```

<br>

Related, we can also infer if a CN change is a gain or loss, relative to ploidy:

```{r}
dlptools::mark_cn_relative_to_ploidy(
  ex_reads, # or can pass segments df
  df_type = "reads" # df_type = "segs"
) |>
  dplyr::relocate(
    # key added columns
    mode_ploidy, cn_v_ploidy
  ) |>
  dplyr::slice_head(n = 5)
```

<br>
<br>

## Long to Wide Reads (or segments)

Some functions require read state information to be in wide format vs long, with cell_ids as rows and chr_start_end as columns, and the states as cells.

```{r}
ex_reads_w <- dlptools::convert_long_reads_to_wide(ex_reads)

ex_reads_w[1:4, 1:4]
```



<br>
<br>
<br>


## Other

Phylogenetic trees made by Stika take some formatting before they can be plotted:

```{r, eval=FALSE}
dlptools::format_sitka_tree()
```

this function drops `locus` tips and removes the `cell_` part of cell id names on tips. This way, the trees can be aligned to cell ids in the heatmaps.
